Neural Circuit Organization of the Ctenophore Apical Organ

### 2025-03-18 manuscript/Manuscript.qmd
edit main text

As mentioned above, following the identification of the presynaptic structures, we reconstructed three major Syncytial neurons. Each of these Syncytial neurons was a cell with multiple nuclei, with membranes fused by continuous plasma membranes. These neurons are distinct in morphology from the Syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously reconstructed in 3D [@Burkhardt2023]. Our findings represent the second documented discovery of Syncytial neurons in ctenophores. Furthermore, these neurons exhibited clear morphological differences when compared to other sensory cells reported in the same study, such as mesogleal neurons and sensory cells with presynaptic structures and cilia (types 1-4)[@Burkhardt2023]. Based on their distinct spatial relationships, we were able to classify these three Syncytial neurons into two categories. The first type is a larger "AO neuron_Q1234," which possesses four (or possibly five?) nuclei spanning four quadrants. It contained X presynaptic structures. The second type consists of "AO neuron_Q12" and "AO neuron_Q34," each having two nuclei spanning two quadrants, and each containing X presynaptic structures.

![**Figure 2. Identification of Three Syncytial Neurons in the Apical Organ**\
(A) 3D reconstruction of the aboral nerve net (ANN) Q1Q2Q3Q4 (Q1-4), which spans all four quadrants and contains multiple nuclei (blue). The spheres represent the positions of individual nuclei. The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(B) 3D reconstruction of two ANN Q1Q2 (pink) and Q3Q4 (orange), each spanning two quadrants and containing multiple nuclei. The spheres indicate the positions of the nuclei. The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(C) Graph showing the number of mitochondria per cell and the proportion of mitochondria involved in synapse formation for each cell type. Green represents mitochondria associated with synapses, while gray represents mitochondria not involved in synapse formation. ANN has a significantly higher number of mitochondria compared to other cell types, with most of them forming synapses. The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
(D) Localization of mitochondria within AO neuron Q1Q2Q3Q4. Red indicates mitochondria associated with the presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.\
(E) Representative electron micrographs of presynaptic triad structures observed in the dataset and their schematic diagrams. The left diagram illustrates synaptic projections from the center ANN (ANN Q1-4) to the lateral ANN (ANN Q3Q4), while the right diagram shows synaptic projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN Q1-4). The abbreviations are as follows: mi (mitochondrion), er (endoplasmic reticulum), and sv (synaptic vesicles).\
(F) Plot showing the three-dimensional positions of synapses. Synapses from the center ANN to the lateral ANN are shown in magenta, while synapses from the lateral ANN to the center ANN are shown in cyan. Blue dots indicate the locations of autapses within ANN Q1-4.\
(G) Neural circuit map of ANN connectivity. The blue, orange, and pink circles represent individual neurons in ANN Q1-4, Q2Q3, and Q3Q4, respectively. Magenta arrows indicate synaptic connections from the center ANN to the lateral ANN, while cyan arrows indicate connections from the lateral ANN to the center ANN. Numbers denote the number of synapses. Blue arrows represent autapses in ANN Q1-4 along with their synapse counts.](figures/Figure2.png)

@@

We discovered that AO neurons form synaptic connections with each other, and that AO neuron_Q1234 forms self-synapses, or “autapses.” To our knowledge, previous reports have not identified synaptic connections between subepithelial nerve net (SNN) neurons. Thus, our results represent the first report of synaptic connectivity between neurons in the apical organ, forming a network in ctenophores (subject to confirmation). These AO neuron networks were found to form many presynaptic structures in relation to the gravity-sensing balancer cells.

@@

From AO neuron_Q1Q2, synapses were formed with 6 of the 37 balancer cells in the Q1 region, and 8 of the 32 balancer cells in the Q2 region. Similarly, from AO neuron_Q3Q4, synapses were formed with 1 of the 32 balancer cells in the Q3 region, and 5 of the 28 balancer cells in the Q4 region. AO neuron_Q1234 formed input synapses with balancer cells in the Q1 region (7 cells), Q2 region (11 cells), Q3 region (6 cells), and Q4 region (10 cells). Some balancer cells received inputs from both AO neuron_Q1Q2 or AO neuron_Q3Q4 and AO neuron_Q1234. While previous studies have suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez-Nicaise1974], our data did not reveal any synaptic inputs from balancer cells to AO neurons. However, we identified "bridge cells" that actively formed input synapses with AO neurons, highlighting their potential role in the network.

@@

From our data, we identified several cell groups that form multiple afferent synapses with AO neurons. These cells, based on their morphological features, were found to be the bridge cells first described by Tamm et al. in 2002 [@Tamm_2002]. These cells are characterized by bundles of elongated processes filled with microtubules that arch over the epithelial layer, resembling a bridge. They originate from the base of paired balancer cells along the tentacle surface and extend across the sagittal plane toward the base of the opposite balancer cells. In regions where the mitochondria of the bridge cells are localized (approximately 30%), a presynaptic triad structure, similar to that of AO neurons, was found, containing synaptic vesicles and smooth endoplasmic reticulum.

Our three-dimensional reconstruction data revealed that the bridge cells form two distinct cell groups across the sagittal plane, between the Q1Q2 and Q3Q4 regions. In the Q1Q2 region, 14 cells were identified, while in the Q3Q4 region, 12 cells were identified, totaling 26 bridge cells. Nearly all of these bridge cells (25 out of 26) exhibited afferent synapses from AO neurons. For bridge cells located in the Q1Q2 region, synaptic inputs came primarily from AO neuron_Q1Q2 (11 cells), AO neuron_Q1234 (1 cell), or both (2 cells). For bridge cells located in the Q3Q4 region, synaptic inputs were received from AO neuron_Q3Q4 (1 cell), AO neuron_Q1234 (7 cells), or both (1 cell). In other words, bridge cells in the Q1Q2 region mainly received input from AO neuron_Q1Q2, while those in the Q3Q4 region received input primarily from AO neuron_Q1234, showing a distinct pattern of input in the two regions.

Some bridge cells also formed afferent synapses with AO neurons. For example, bridge cells in the Q1Q2 region formed synapses with AO neuron_Q1Q2 (3 cells) or both AO neuron_Q1Q2 and AO neuron_Q1234 (2 cells). Bridge cells in the Q3Q4 region formed synapses with AO neuron_Q3Q4 (1 cell) or AO neuron_Q1234 (6 cells). A notable difference was observed between the Q1Q2 and Q3Q4 regions in the proportion of synaptic inputs from bridge cells to AO neurons.

Interestingly, in both regions, bridge cells also formed synapses with adjacent bridge cells. However, no synaptic input was found from bridge cells across the sagittal plane to those in the opposite region. To analyze the grouped synaptic connectivity graph, we classified cell types within each region, collapsed cells of the same type into a single node, summed the number of synapses, and explored directed pathways to effectors (balancer cell groups). The results revealed a feedback pathway from AO neurons through the synaptic connections formed by bridge cells, thereby shedding light on the neural circuit structure involving these cells.

![**Figure 3. Reconstruction of the balancers and bridges that receives synapses from the ANN.**\
(A) Graph showing the number of synaptic outputs from the ANN circuit for each cell type. The balancer and bridge cell types receive relatively high outputs from the ANN circuit. ANN represents the number of synapses within the ANN circuit. The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
(B) Connectivity map of three AO neurons (Q1234 in blue, Q12 in pink, and Q34 in orange) and balancer ciliated cells (light blue). The thickness of the arrows and the numbers correspond to the number of synapses. Synaptic structures from the same neuron targeting the same balancer cell are grouped into hexagons, with the number of cells added to the label.\
(C) 3D reconstruction of bridge cells spanning the Q1Q2 and Q3Q4 quadrants. The Q1Q2-side bridge cells (8 cells) are shown in blue, while the Q3Q4-side bridge cells (6 cells) are shown in xxx color. The morphology of individual bridge cells extending across opposite quadrant regions is depicted. The spheres represent the positions of individual nuclei. The left panel shows an aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(D) Mitochondrial localization within bridge cells and associated presynaptic triad structures. Red indicates mitochondria associated with presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified. The left panel shows a dorsal view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(E) Synaptic connections from AO neurons to balancer ciliated cells. The left panel presents an aboral view, while the right panel shows a lateral view and a tentacular plane view. Balancer ciliated cells are depicted in light gray, AO neuron Q1234 in black, AO neuron Q12 in orange, and AO neuron Q34 in blue.\
(F) Synaptic connections between AO neurons and bridge cells. The positions of synapses from AO neurons to bridge cells (red) and from bridge cells to AO neurons (light blue) are indicated. The left panel shows an aboral view, while the right panel presents a tentacular plane lateral view. Bridge cells are shown in light gray, AO neurons in Q1234 are in black, AO neurons in Q12 are in orange, and AO neurons in Q34 are in blue.\
(G) Connectivity matrix of the gravity-sensing neural circuit. Columns represent presynaptic cell groups, while rows represent postsynaptic cell groups. The numbers and varying shades of blue correspond to the number of synapses.\
(H) Complete connectivity map of the gravity-sensing neural circuit. Cells belonging to the same group are enclosed in hexagons, and the number of cells is added to their labels. The thickness of the arrows and the numerical values indicate the number of synapses. AO neurons in Q1234 are shown in black, AO neurons in Q12 in yellow, AO neurons in Q34 in blue, balancer cell groups in gray, and bridge cell groups in pink.](figures/Figure3.png)

@@

To investigate the function of gravity-sensitive circuits, we conducted a comparative analysis of balancer ciliary movements across different regions. Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. Furthermore, it has been suggested that differences in the asymmetry of statolith morphology and the elongation direction of balancer cilia between the tentacular and sagittal planes could result in variations in the force received from the statolith [@Tamm_2014; @Tamm2015]. Building on these findings, we standardized conditions for observing balancer cilia by tilting the microscope by 90 degrees and fixing samples such that the aboral-oral axis was aligned parallel to the vertical stage, with the aboral side facing upward. Using connectome analysis, we aimed to evaluate the neural circuit differences by analyzing the control patterns of left and right balancer cilia in the tentacular and sagittal planes. Specifically, we hypothesized that differences in control between Q1 (Q2) and Q3 (Q4) regions, as well as between Q1 (Q4) and Q2 (Q3) regions, would reveal distinct functions of AO neurons (e.g., AO neuron Q12/Q34 vs. AO neuron Q1234). For the sagittal plane, we recorded balancer ciliary movements in the Q1(4) and Q2(3) regions (left and right sides of the image) and, for the tentacular plane, in the Q1(3) and Q2(4) regions, using a high-speed camera (100 fps) for 2 minutes (12,000 frames). Regions of interest (ROIs) were defined in areas where brightness changes indicated ciliary beating. We measured these brightness changes to generate dynamics graphs of balancer ciliary movements for both planes.

During the 2-minute recordings, balancer cilia exhibited movements that could be fast, slow, or stop abruptly (arrest) and start moving again (re-beat). Occasionally, large contraction responses caused the entire cydippid to move out of frame, and data from these moments were excluded. We focused on two metrics for left-right comparison: (1) the timing of ciliary arrest and (2) the timing of re-beat initiation following an arrest. Due to the low baseline beating frequency of the cilia, rapid synaptic input responses were not evaluated, suggesting that ciliary beat regulation may involve non-synaptic control mechanisms. The analysis revealed that both arrest and re-beat events were synchronized between left and right balancers in both planes within a few milliseconds, suggesting that these events are governed by synaptic input from neurons. Interestingly, while the re-beat timing showed no significant differences between the sagittal and tentacular planes, arrest timing exhibited greater variability in the tentacular plane. Comparing these findings with neural circuit diagrams suggests that in the sagittal plane, the same AO neurons (Q12 or Q34) project to both balancer cell groups, transmitting signals simultaneously and enabling synchronization. In contrast, in the tentacular plane, separate AO neurons project to different regions, potentially introducing an additional synaptic step, which may account for the observed variability in arrest timing.

@@

Larvae of Mnemiopsis leidyi (5 days old) were cryofixed using a high-pressure freezing apparatus (BAL-TEC HPM 010, Balzers) and immediately transferred to liquid nitrogen for storage. The frozen samples were processed in a substitution medium containing 2% (w/v) osmium tetroxide and 0.5% uranyl acetate in acetone, using a cryo-substitution device (EM AFS-2, Leica). Cryo-substitution was performed by gradually raising the temperature, and the samples were embedded in epoxy resin. Serial sections of 50 nm thickness were prepared using a Reichert Jung Ultracut E ultramicrotome and a 45º DiATOME diamond knife. To enhance section adhesion and improve hydrophilicity, a conductive indium tin oxide-coated glass slide (ITO Glass, UQG Optics) was treated with air glow discharge using the PELCO easiGlow system (Ted Pella, Inc.), rendering the carbon film surface negatively charged. Section ribbons were collected on the prepared glass slides, slowly dried to ensure proper stretching, and firmly adhered to the glass surface. The sections on glass were stained with UranyLess and lead citrate (Reynolds) using airless staining bottles (Delta Microscopies). The glass slides were mounted on STEM-specific stages (Zeiss) using Copper Foil EMI Shielding Tape (3M). Imaging was performed using a Gemini SEM 500 (Zeiss) equipped with SmartSEM and Atlas 5 imaging software (Zeiss). The full dataset consisted of 620 serial sections, each 50 nm thick. The imaging resolution was 2.8 nm/pixel, using electron holography transmission at an acceleration voltage of 1.5 kV (in-lens detector, dwell time: 3 µs).

@@

The reconstruction of cells and connectivity was performed using CATMAID [@Saalfeld2009], analyzing a total of 620 serial sections. To digitally reconstruct all neurons within the serial TEM dataset, we utilized the collaborative web-based application CATMAID, installed on a local server [@Saalfeld2009; @Schneider-Mizell2016]. To mark the locations of cell bodies, we placed tags at the approximate center of each nucleus within the dataset. At each nuclear center, the radius of the single node was adjusted according to the size of the cell body in that specific layer. All skeletons were rooted at their respective cell bodies, and nodes were tagged as "soma." Synapses were identified based on four key structural features: the cell membrane, synaptic vesicles, the endoplasmic reticulum, and mitochondria. Most synapses could be verified across consecutive sections, ensuring accurate annotation and connectivity mapping.

@@

Each cell is assigned a specific type name based on its category (balancer, biC, bridge, bristle, dense vesicle cell, dome, EF, IcMC, LB, lithocyte, monoC, non-C, plumose cell, SNN, ssCG, ANN, stCG), resulting in a total of 17 cell types. Additionally, a quadrant number is appended to the cell type name following an underscore ("\_") to indicate the cell body’s location. For example, cells located in the first quadrant are annotated with "Q1." If a cell is situated between the first and second quadrants, the annotation "Q1Q2" is applied.

@@

We conducted the analysis using the cydippid stage of *Mnemiopsis leidyi* at five days post-fertilization. A coverslip with a thin layer of Vaseline applied to two edges was lightly placed on a slide glass. Filtered natural seawater was introduced under the coverslip, along with the cydippids. By gently moving the coverslip, the desired orientation for observation was adjusted, and once positioned, slight pressure was applied to immobilize the cydippids. To stabilize the statolith's position, the microscope was tilted 90 degrees, arranging the stage vertically. The movement of the balancer cilia was observed using a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a highly sensitive CMOS monochrome camera optimized for the near-infrared (NIR) range (UI-3360CP-NIR-GL Rev.2, iDS). Observations were made using a 40x objective lens (Objective LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27) with glycerine immersion. For recording, we employed a custom-made NIR LED strobe illumination system (wavelength: 850 nm) synchronized with the camera's exposure signals. The camera and LED strobe system were operated using the Video Capture Software BohNavi. Images were recorded at a resolution of 640×480 pixels with a frame rate of 100 fps for 2 minutes, using 0.05 ms pulses from the LED. To analyze the ciliary beating of the balancer cilia, we utilized the Multi Kymograph project in Fiji software.

>>>>

As mentioned above, following the identification of the presynaptic structures, we reconstructed three major Syncytial neurons. Each of these Syncytial neurons was a cell with multiple nuclei, with membranes fused by continuous plasma membranes. These neurons are distinct in morphology from the Syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously reconstructed in 3D [@Burkhardt2023]. Our findings represent the second documented discovery of Syncytial neurons in ctenophores. Furthermore, these neurons exhibited clear morphological differences when compared to other sensory cells reported in the same study, such as mesogleal neurons and sensory cells with presynaptic structures and cilia (types 1-4)[@Burkhardt2023]. Based on their distinct spatial relationships, we were able to classify these three Syncytial neurons into two categories. The first type is a larger "ANN_Q1-4," which possesses four (or possibly five?) nuclei spanning four quadrants. It contained X presynaptic structures. The second type consists of "ANN_Q1Q2" and "ANN_Q3Q4," each having two nuclei spanning two quadrants, and each containing X presynaptic structures.

![**Figure 2. Identification of Three Syncytial Neurons in the Apical Organ**\
(A) 3D reconstruction of the aboral nerve net (ANN) Q1Q2Q3Q4 (Q1-4), which spans all four quadrants and contains multiple nuclei (blue). The spheres represent the positions of individual nuclei. The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(B) 3D reconstruction of two ANN Q1Q2 (pink) and Q3Q4 (orange), each spanning two quadrants and containing multiple nuclei. The spheres indicate the positions of the nuclei. The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(C) Graph showing the number of mitochondria per cell and the proportion of mitochondria involved in synapse formation for each cell type. Green represents mitochondria associated with synapses, while gray represents mitochondria not involved in synapse formation. ANN has a significantly higher number of mitochondria compared to other cell types, with most of them forming synapses. The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
(D) Localization of mitochondria within ANN Q1-4. Red indicates mitochondria associated with the presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.\
(E) Representative electron micrographs of presynaptic triad structures observed in the dataset and their schematic diagrams. The left diagram illustrates synaptic projections from the center ANN (ANN Q1-4) to the lateral ANN (ANN Q3Q4), while the right diagram shows synaptic projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN Q1-4). The abbreviations are as follows: mi (mitochondrion), er (endoplasmic reticulum), and sv (synaptic vesicles).\
(F) Plot showing the three-dimensional positions of synapses. Synapses from the center ANN to the lateral ANN are shown in magenta, while synapses from the lateral ANN to the center ANN are shown in cyan. Blue dots indicate the locations of autapses within ANN Q1-4.\
(G) Neural circuit map of ANN connectivity. The blue, orange, and pink circles represent individual neurons in ANN Q1-4, Q2Q3, and Q3Q4, respectively. Magenta arrows indicate synaptic connections from the center ANN to the lateral ANN, while cyan arrows indicate connections from the lateral ANN to the center ANN. Numbers denote the number of synapses. Blue arrows represent autapses in ANN Q1-4 along with their synapse counts.](figures/Figure2.png)

@@

We discovered that ANN neurons form synaptic connections with each other, and that ANN_Q1-4 forms self-synapses, or “autapses.” To our knowledge, previous reports have not identified synaptic connections between subepithelial nerve net (SNN) neurons. Thus, our results represent the first report of synaptic connectivity between neurons in the apical organ, forming a network in ctenophores (subject to confirmation). These ANN networks were found to form many presynaptic structures in relation to the gravity-sensing balancer cells.

@@

From ANN_Q1Q2, synapses were formed with 6 of the 37 balancer cells in the Q1 region, and 8 of the 32 balancer cells in the Q2 region. Similarly, from ANN_Q3Q4, synapses were formed with 1 of the 32 balancer cells in the Q3 region, and 5 of the 28 balancer cells in the Q4 region. ANN_Q1-4 formed input synapses with balancer cells in the Q1 region (7 cells), Q2 region (11 cells), Q3 region (6 cells), and Q4 region (10 cells). Some balancer cells received inputs from both ANN_Q1Q2 or ANN_Q3Q4 and ANN_Q1-4. While previous studies have suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez-Nicaise1974], our data did not reveal any synaptic inputs from balancer cells to ANN neurons. However, we identified "bridge cells" that actively formed input synapses with ANN neurons, highlighting their potential role in the network.

@@

From our data, we identified several cell groups that form multiple afferent synapses with ANNs. These cells, based on their morphological features, were found to be the bridge cells first described by Tamm et al. in 2002 [@Tamm_2002]. These cells are characterized by bundles of elongated processes filled with microtubules that arch over the epithelial layer, resembling a bridge. They originate from the base of paired balancer cells along the tentacle surface and extend across the sagittal plane toward the base of the opposite balancer cells. In regions where the mitochondria of the bridge cells are localized (approximately 30%), a presynaptic triad structure, similar to that of ANNs, was found, containing synaptic vesicles and smooth endoplasmic reticulum.

Our three-dimensional reconstruction data revealed that the bridge cells form two distinct cell groups across the sagittal plane, between the Q1Q2 and Q3Q4 regions. In the Q1Q2 region, 14 cells were identified, while in the Q3Q4 region, 12 cells were identified, totaling 26 bridge cells. Nearly all of these bridge cells (25 out of 26) exhibited afferent synapses from ANNs. For bridge cells located in the Q1Q2 region, synaptic inputs came primarily from ANN_Q1Q2 (11 cells), ANN_Q1-4 (1 cell), or both (2 cells). For bridge cells located in the Q3Q4 region, synaptic inputs were received from ANN_Q3Q4 (1 cell), ANN_Q1-4 (7 cells), or both (1 cell). In other words, bridge cells in the Q1Q2 region mainly received input from ANN_Q1Q2, while those in the Q3Q4 region received input primarily from ANN_Q1-4, showing a distinct pattern of input in the two regions.

Some bridge cells also formed afferent synapses with ANN neurons. For example, bridge cells in the Q1Q2 region (3 cells) formed synapses with ANN_Q1Q2 or (2 cells) both ANN_Q1Q2 and ANN_Q1-4. Bridge cells in the Q3Q4 region formed synapses with ANN_Q3Q4 (1 cell) or ANN_Q1-4 (6 cells). A notable difference was observed between the Q1Q2 and Q3Q4 regions in the proportion of synaptic inputs from bridge cells to ANN neurons.

Interestingly, in both regions, bridge cells also formed synapses with adjacent bridge cells. However, no synaptic input was found from bridge cells across the sagittal plane to those in the opposite region. To analyze the grouped synaptic connectivity graph, we classified cell types within each region, collapsed cells of the same type into a single node, summed the number of synapses, and explored directed pathways to effectors (balancer cell groups). The results revealed a feedback pathway from ANN neurons through the synaptic connections formed by bridge cells, thereby shedding light on the neural circuit structure involving these cells.

![**Figure 3. Reconstruction of the balancers and bridges that receives synapses from the ANN.**\
(A) Graph showing the number of synaptic outputs from the ANN circuit for each cell type. The balancer and bridge cell types receive relatively high outputs from the ANN circuit. ANN represents the number of synapses within the ANN circuit. The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
(B) Connectivity map of three ANN neurons (Q1-4 in blue, Q1Q2 in pink, and Q3Q4 in orange) and balancer ciliated cells (light blue). The thickness of the arrows and the numbers correspond to the number of synapses. Synaptic structures from the same neuron targeting the same balancer cell are grouped into hexagons, with the number of cells added to the label.\
(C) 3D reconstruction of bridge cells spanning the Q1Q2 and Q3Q4 quadrants. The Q1Q2-side bridge cells (8 cells) are shown in blue, while the Q3Q4-side bridge cells (6 cells) are shown in xxx color. The morphology of individual bridge cells extending across opposite quadrant regions is depicted. The spheres represent the positions of individual nuclei. The left panel shows an aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(D) Mitochondrial localization within bridge cells and associated presynaptic triad structures. Red indicates mitochondria associated with presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified. The left panel shows a dorsal view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(E) Synaptic connections from ANN neurons to balancer ciliated cells. The positions of synapses from ANNs to balancers (magenta) are indicated. The left panel presents an aboral view, while the right panel shows a lateral view and a tentacular plane view. Balancer ciliated cells are depicted in light gray.\
(F) Synaptic connections between ANNs and bridge cells. The positions of synapses from ANNs to bridge cells (magenta) and from bridge cells to ANNs (light blue) are indicated. The left panel shows an aboral view, while the right panel presents a tentacular plane lateral view. Bridge cells are shown in light gray.\
(G) Connectivity matrix of the gravity-sensing neural circuit. Columns represent presynaptic cell groups, while rows represent postsynaptic cell groups. The numbers and varying shades of blue correspond to the number of synapses.\
(H) Complete connectivity map of the gravity-sensing neural circuit. Cells belonging to the same group are enclosed in hexagons, and the number of cells is added to their labels. The thickness of the arrows and the numerical values indicate the number of synapses. ANN Q1-4 are shown in blue, ANN Q1Q2 in orange, ANNs in Q3Q4 in pink, balancer cell groups in gray, and bridge cell groups in yellow.](figures/Figure3.png)

@@

To investigate the function of gravity-sensitive circuits, we conducted a comparative analysis of balancer ciliary movements across different regions. Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. Furthermore, it has been suggested that differences in the asymmetry of statolith morphology and the elongation direction of balancer cilia between the tentacular and sagittal planes could result in variations in the force received from the statolith [@Tamm_2014; @Tamm2015]. Building on these findings, we standardized conditions for observing balancer cilia by tilting the microscope by 90 degrees and fixing samples such that the aboral-oral axis was aligned parallel to the vertical stage, with the aboral side facing upward. Using connectome analysis, we aimed to evaluate the neural circuit differences by analyzing the control patterns of left and right balancer cilia in the tentacular and sagittal planes. Specifically, we hypothesized that differences in control between Q1 (Q2) and Q3 (Q4) regions, as well as between Q1 (Q4) and Q2 (Q3) regions, would reveal distinct functions of ANNs (e.g., ANN Q1Q2/Q3Q4 vs. ANN Q1-4). For the sagittal plane, we recorded balancer ciliary movements in the Q1(4) and Q2(3) regions (left and right sides of the image) and, for the tentacular plane, in the Q1(3) and Q2(4) regions, using a high-speed camera (100 fps) for 2 minutes (12,000 frames). Regions of interest (ROIs) were defined in areas where brightness changes indicated ciliary beating. We measured these brightness changes to generate dynamics graphs of balancer ciliary movements for both planes.

During the 2-minute recordings, balancer cilia exhibited movements that could be fast, slow, or stop abruptly (arrest) and start moving again (re-beat). Occasionally, large contraction responses caused the entire cydippid to move out of frame, and data from these moments were excluded. We focused on two metrics for left-right comparison: (1) the timing of ciliary arrest and (2) the timing of re-beat initiation following an arrest. Due to the low baseline beating frequency of the cilia, rapid synaptic input responses were not evaluated, suggesting that ciliary beat regulation may involve non-synaptic control mechanisms. The analysis revealed that both arrest and re-beat events were synchronized between left and right balancers in both planes within a few milliseconds, suggesting that these events are governed by synaptic input from neurons. Interestingly, while the re-beat timing showed no significant differences between the sagittal and tentacular planes, arrest timing exhibited greater variability in the tentacular plane. Comparing these findings with neural circuit diagrams suggests that in the sagittal plane, the same ANNs (Q1Q2 or Q3Q4) project to both balancer cell groups, transmitting signals simultaneously and enabling synchronization. In contrast, in the tentacular plane, separate ANNs project to different regions, potentially introducing an additional synaptic step, which may account for the observed variability in arrest timing.

@@

Larvae of *Mnemiopsis leidyi* (5 days old) were cryofixed using a high-pressure freezing apparatus (BAL-TEC HPM 010, Balzers) and immediately transferred to liquid nitrogen for storage. The frozen samples were processed in a substitution medium containing 2% (w/v) osmium tetroxide and 0.5% uranyl acetate in acetone, using a cryo-substitution device (EM AFS-2, Leica). Cryo-substitution was performed by gradually raising the temperature, and the samples were embedded in epoxy resin. Serial sections of 50 nm thickness were prepared using a Reichert Jung Ultracut E ultramicrotome and a 45º DiATOME diamond knife. To enhance section adhesion and improve hydrophilicity, a conductive indium tin oxide-coated glass slide (ITO Glass, UQG Optics) was treated with air glow discharge using the PELCO easiGlow system (Ted Pella, Inc.), rendering the carbon film surface negatively charged. Section ribbons were collected on the prepared glass slides, slowly dried to ensure proper stretching, and firmly adhered to the glass surface. The sections on glass were stained with UranyLess and lead citrate (Reynolds) using airless staining bottles (Delta Microscopies). The glass slides were mounted on STEM-specific stages (Zeiss) using Copper Foil EMI Shielding Tape (3M). Imaging was performed using a Gemini SEM 500 (Zeiss) equipped with SmartSEM and Atlas 5 imaging software (Zeiss). The full dataset consisted of 620 serial sections, each 50 nm thick. The imaging resolution was 2.8 nm/pixel, using electron holography transmission at an acceleration voltage of 1.5 kV (in-lens detector, dwell time: 3 µs).

@@

The reconstruction of cells and connectivity was performed using CATMAID [@Saalfeld2009], analyzing a total of 619 serial sections. To digitally reconstruct all neurons within the serial TEM dataset, we utilized the collaborative web-based application CATMAID, installed on a local server [@Saalfeld2009; @Schneider-Mizell2016]. To mark the locations of cell bodies, we placed tags at the approximate center of each nucleus within the dataset. At each nuclear center, the radius of the single node was adjusted according to the size of the cell body in that specific layer. All skeletons were rooted at their respective cell bodies, and nodes were tagged as "soma." Synapses were identified based on four key structural features: the cell membrane, synaptic vesicles, the endoplasmic reticulum, and mitochondria. Most synapses could be verified across consecutive sections, ensuring accurate annotation and connectivity mapping.

@@

Each cell is assigned a specific type name based on its category (balancer, bridge, bristle, ciliated grooves, dense vesicle cell, dome, epithelial floor (ef), intracellular multiciliated cells (imc), lamellate bodies (lb), lithocyte, plumose cell, ANN) resulting in a total of 12 cell types, or those that could not be classified: presence and number of cilia (biciliated (biC), monociliated (monoC), multiciliated (multiC), non-ciliated (non-C)) in 4 ciliated cell types. Additionally, a quadrant number is appended to the cell type name following an underscore ("\_") to indicate the cell body’s location. For example, cells located in the first quadrant are annotated with "Q1." If a cell is situated between the first and second quadrants, the annotation "Q1Q2" is applied.

@@

We conducted the analysis using the cydippid stage of *M. leidyi* at five days post-fertilization. A coverslip with a thin layer of Vaseline applied to two edges was lightly placed on a slide glass. Filtered natural seawater was introduced under the coverslip, along with the cydippids. By gently moving the coverslip, the desired orientation for observation was adjusted, and once positioned, slight pressure was applied to immobilize the cydippids. To stabilize the statolith's position, the microscope was tilted 90 degrees, arranging the stage vertically. The movement of the balancer cilia was observed using a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a highly sensitive CMOS monochrome camera optimized for the near-infrared (NIR) range (UI-3360CP-NIR-GL Rev.2, iDS). Observations were made using a 40x objective lens (Objective LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27) with glycerine immersion. For recording, we employed a custom-made NIR LED strobe illumination system (wavelength: 850 nm) synchronized with the camera's exposure signals. The camera and LED strobe system were operated using the Video Capture Software BohNavi. Images were recorded at a resolution of 640×480 pixels with a frame rate of 100 fps for 2 minutes, using 0.05 ms pulses from the LED. To analyze the ciliary beating of the balancer cilia, we utilized the Multi Kymograph project in Fiji software.
